---
title: "Assignment3"
author: "RF"
date: "3/6/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Multivariate linear models

In this assignment you will investigate the relation between different aspects of IQ and symptom severity in ASD (as measured by ADOS). The assignment is meant to make you practice on linear models, multiple predictors and interactions.

The data you will use is the outcome of the psychological testing of the children you analyzed in methods 3.
Data: https://www.dropbox.com/s/hom2qnmp9hyfwmu/Assignment3Data.csv?dl=0
The data consists of ChildID, gender, age, diagnosis (ASD), symptom severity (ADOS), several aspects of IQ (NonVerbal/Perceptual,Verbal,Social) and the visit at which the tests were administered. The questions should be answered by focusing on visit 1.
```{r}
data <- read.csv("Assignment3Data.csv")
library(rethinking)
```

The questions you will have to answer are:

1. Assess the relation between symptom severity and IQ (focus on visit 1 and children with ASD) and report the model, a plot of the model, a couple of lines describing the quality of the model and interpreting the results. 
P.S. Should you scale? 
P.P.S. Don't forget to motivate your priors. 
P.P.P.S. At least one plot for results and a plot for quality of each model (here and in the next questions) would be appreciated.
1.1. Verbal IQ and ADOS
1.2. Non Verbal IQ and ADOS
1.3. Social IQ and ADOS
```{r}

d1 = data.frame(subset(data, c(Visit == 1 & ASD == 1)))
dens(d1$ADOS)


#as we don't know the mean of the distribution we use the mean of the outcome, as the best suggestion
mean(d1$ADOS)

flist = alist(
  ADOS ~ dnorm(mu, sigma), # ADOS is the ASD severity, what we are interested in
  mu ~ dnorm(14,3), # 14 is the mean, a SD of 3 allows 3 sd's on each side of the mean within the data
  sigma ~ dunif(0,21)) # 0 is the least possible error, and 21 is the biggest possible error of ADOS

# fitting the model to the data
m1 = map(flist, data = d1)

#inspecting MAP
precis(m1)


# 1.1 model for verbal IQ and ADOS
m2= map(
      alist(
          ADOS ~ dnorm(mu, sigma), 
                mu <- a + b*VerbalIQ, 
                a ~ dnorm(14,3), 
                b ~ dnorm(0,25), 
                sigma ~ dunif(0,21)
          ), 
          data = d1)

precis(m2, corr=T)

#sampling from the posterior
post <- extract.samples( m2 , n=1e4 )
VerbalIQ.seq <- seq( from=8 , to=33 , by=1 )

sim.ADOS = sim(m2, data =list(VerbalIQ = VerbalIQ.seq)) # simulating data for verbal IQ
ADOS.PI = apply(sim.ADOS,2,PI, prob =0.89) # making percentile intervals

# plotting raw data
plot( ADOS ~ VerbalIQ , data=d1 , col=col.alpha(rangi2,0.5) )
# plot the MAP line, aka the mean mu for each weight
abline(m2)
# plot a shaded region for 89% HPDI
for ( i in 1:20 )
  abline( a=post$a[i] , b=post$b[i] , col=col.alpha("black",0.3) )
shade( ADOS.PI , VerbalIQ.seq )

# The plot is more overestimating than underestimating, as all of the points above the lines are withing the 89% interval, but not all of the dots below are within the shade

# 1.2 model for non-verbal IQ and ADOS
m3= map(
      alist(
          ADOS ~ dnorm(mu, sigma), 
                mu <- a + b*NonVerbalIQ, 
                a ~ dnorm(14,3), 
                b ~ dnorm(0,29), 
                sigma ~ dunif(0,21)
          ), 
          data = d1)

precis(m3)

#sampling from the posterior
post <- extract.samples( m3 , n=1e4 )
NonVerbalIQ.seq <- seq( from=13 , to=42 , by=1 )

sim.ADOS = sim(m3, data =list(NonVerbalIQ = NonVerbalIQ.seq))
ADOS.PI = apply(sim.ADOS,2,PI, prob =0.89)


plot( ADOS ~ NonVerbalIQ , data=d1 , col=col.alpha(rangi2,0.5) )
# plot the MAP line, aka the mean mu for each weight
abline(m3)
# plot a shaded region for 89% HPDI
for ( i in 1:20 )
  abline( a=post$a[i] , b=post$b[i] , col=col.alpha("black",0.3) )
shade( ADOS.PI , NonVerbalIQ.seq )

# 1.3 model for social IQ and ADOS
m4= map(
      alist(
          ADOS ~ dnorm(mu, sigma), 
                mu <- a + b*SocialIQ, 
                a ~ dnorm(14,3), 
                b ~ dnorm(0,44), 
                sigma ~ dunif(0,21)
          ), 
          data = d1)

precis(m4, corr=T)

#sampling from the posterior
post <- extract.samples( m4 , n=1e4 )
SocialIQ.seq <- seq( from=61 , to=105 , by=1 )

sim.ADOS = sim(m4, data =list(SocialIQ = SocialIQ.seq))
ADOS.PI = apply(sim.ADOS,2,PI, prob =0.89)


plot( ADOS ~ SocialIQ , data=d1 , col=col.alpha(rangi2,0.5) )
# plot the MAP line, aka the mean mu for each weight
abline(m4)
# plot a shaded region for 89% HPDI
for ( i in 1:20 )
  abline( a=post$a[i] , b=post$b[i] , col=col.alpha("black",0.3) )
shade( ADOS.PI , SocialIQ.seq )
```

2. Do the different aspects of IQ account for different portions of the variance in ADOS? 
2.1. Does it make sense to have all IQ measures in the same model? First write a few lines answering the question and motivating your answer, including a discussion as to what happens when you put all of them in the same model. Then build a model following your answer. If your answer is "no", you are not free, you still have to answer: are there alternative ways of answering the question?
2.2. Build the model, assess its quality, write a few lines interpreting the results + plots.

```{r}
m5= map(
      alist(
          ADOS ~ dnorm(mu, sigma), 
                mu <- a + bv*VerbalIQ + bn*NonVerbalIQ + bs*SocialIQ, 
                a ~ dnorm(14,3), 
                bv ~ dnorm(0,25),
                bn ~ dnorm(0,29),
                bs ~ dnorm(0,44),
                sigma ~ dunif(0,21)
          ), 
          data = d1)

precis(m5, digits = 3)

# plotting the impact of changes in Verbal IQ on predictions
# prepare new counterfactual data - non-verbal vs verbal IQ
N.avg = mean(d1$NonVerbalIQ)
S.avg = mean(d1$SocialIQ)
V.seq = seq(from=8, to= 33, length.out = 30)
pred.data = data.frame(
  VerbalIQ = V.seq,
  NonVerbalIQ = N.avg,
  SocialIQ = S.avg
)

#compute counterfactual mean  (mu)
mu = link(m5, data = pred.data)
mu.mean = apply(mu, 2, mean)
mu.PI = apply(mu, 2, PI)

#simulate counterfactual outcomes
V.sim = sim(m5, data = pred.data, n=1e4)
V.PI = apply(V.sim, 2, PI)

#display predictions, hiding raw data with type="n"
plot(ADOS ~ VerbalIQ, data=d1, type="n")
mtext("NonVerbalIQ + Social = 0")
lines(V.seq, mu.mean)
shade(mu.PI, V.seq)
shade(V.PI, V.seq)

# plotting the impact of changes in NonVerbal IQ on predictions
# prepare new counterfactual data 
V.avg = mean(d1$VerbalIQ)
S.avg = mean(d1$SocialIQ)
N.seq = seq(from=13, to= 42, length.out = 30)
pred.data = data.frame(
  NonVerbalIQ = N.seq,
  VerbalIQ = V.avg,
  SocialIQ = S.avg
)

#compute counterfactual mean divorce (mu)
mu = link(m5, data = pred.data)
mu.mean = apply(mu, 2, mean)
mu.PI = apply(mu, 2, PI)

#simulate counterfactual divorce outcomes
N.sim = sim(m5, data = pred.data, n=1e4)
N.PI = apply(N.sim, 2, PI)

#display predictions, hiding raw data with type="n"
plot(ADOS ~ NonVerbalIQ, data=d1, type="n")
mtext("SocialIQ + VerbalIQ = 0")
lines(N.seq, mu.mean)
shade(mu.PI, N.seq)
shade(N.PI, N.seq)

# plotting the impact of changes in Social IQ on predictions
# prepare new counterfactual data 
V.avg = mean(d1$VerbalIQ)
N.avg = mean(d1$NonVerbalIQ)
S.seq = seq(from=61, to= 105, length.out = 30)
pred.data = data.frame(
  SocialIQ = S.seq,
  VerbalIQ = V.avg,
  NonVerbalIQ = N.avg
)

#compute counterfactual mean divorce (mu)
mu = link(m5, data = pred.data)
mu.mean = apply(mu, 2, mean)
mu.PI = apply(mu, 2, PI)

#simulate counterfactual divorce outcomes
S.sim = sim(m5, data = pred.data, n=1e4)
S.PI = apply(S.sim, 2, PI)

#display predictions, hiding raw data with type="n"
plot(ADOS ~ SocialIQ, data=d1, type="n")
mtext("NonVerbalIQ + VerbalIQ = 0")
lines(S.seq, mu.mean)
shade(mu.PI, S.seq)
shade(S.PI, S.seq)

#dens plot simulated data from model on raw data
sim.m5 = sim(m5, data = d1, n=1e4)
dens(sim.m5, col = "red", xlim = c(-5, 30), ylim = c(0,0.12),  xlab = "ADOS")
par(new=TRUE)
dens(d1$ADOS, xlim = c(-5, 30), ylim = c(0,0.12), xlab = "ADOS")
title("Model 5")
```

3. Let's now include also the TD children. Does it make sense to ask whether IQ and ADOS are related? Motivate your answer. In any case, if you wanted to build a model to answer that question, which model would you build? Run the model, assess its quality, write a few lines interpreting the results.
```{r}
# subsetting all 1st visit
d2=subset(data, Visit == 1)
d2=d2[complete.cases(d2),]
# fitting 
m6= map(
      alist(
          ADOS ~ dnorm(mu, sigma), 
                mu <- a + bv*VerbalIQ + bn*NonVerbalIQ + bs*SocialIQ, 
                a ~ dnorm(7,3), 
                bv ~ dnorm(0,25),
                bn ~ dnorm(0,29),
                bs ~ dnorm(0,54),
                sigma ~ dunif(0,21)
          ), 
          data = d2)

precis(m6, digits = 3, corr = T)

# plotting the impact of changes in Verbal IQ on predictions
# prepare new counterfactual data - non-verbal vs verbal IQ
N.avg = mean(d2$NonVerbalIQ)
S.avg = mean(d2$SocialIQ)
V.seq = seq(from=8, to= 33, length.out = 30)
pred.data = data.frame(
  VerbalIQ = V.seq,
  NonVerbalIQ = N.avg,
  SocialIQ = S.avg
)

#compute counterfactual mean  (mu)
mu = link(m6, data = pred.data)
mu.mean = apply(mu, 2, mean)
mu.PI = apply(mu, 2, PI)

#simulate counterfactual outcomes
V.sim = sim(m6, data = pred.data, n=1e4)
V.PI = apply(V.sim, 2, PI)

#display predictions, hiding raw data with type="n"
plot(ADOS ~ VerbalIQ, data=d2, type="n")
mtext("NonVerbalIQ + Social = 0")
lines(V.seq, mu.mean)
shade(mu.PI, V.seq)
shade(V.PI, V.seq)

# plotting the impact of changes in NonVerbal IQ on predictions
# prepare new counterfactual data 
V.avg = mean(d2$VerbalIQ)
S.avg = mean(d2$SocialIQ)
N.seq = seq(from=13, to= 42, length.out = 30)
pred.data = data.frame(
  NonVerbalIQ = N.seq,
  VerbalIQ = V.avg,
  SocialIQ = S.avg
)

#compute counterfactual mean divorce (mu)
mu = link(m6, data = pred.data)
mu.mean = apply(mu, 2, mean)
mu.PI = apply(mu, 2, PI)

#simulate counterfactual divorce outcomes
N.sim = sim(m6, data = pred.data, n=1e4)
N.PI = apply(N.sim, 2, PI)

#display predictions, hiding raw data with type="n"
plot(ADOS ~ NonVerbalIQ, data=d2, type="n")
mtext("SocialIQ + VerbalIQ = 0")
lines(N.seq, mu.mean)
shade(mu.PI, N.seq)
shade(N.PI, N.seq)

# plotting the impact of changes in Social IQ on predictions
# prepare new counterfactual data 
V.avg = mean(d2$VerbalIQ)
N.avg = mean(d2$NonVerbalIQ)
S.seq = seq(from=61, to= 115, length.out = 30)
pred.data = data.frame(
  SocialIQ = S.seq,
  VerbalIQ = V.avg,
  NonVerbalIQ = N.avg
)

#compute counterfactual mean divorce (mu)
mu = link(m6, data = pred.data)
mu.mean = apply(mu, 2, mean)
mu.PI = apply(mu, 2, PI)

#simulate counterfactual divorce outcomes
S.sim = sim(m6, data = pred.data, n=1e4)
S.PI = apply(S.sim, 2, PI)

#display predictions, hiding raw data with type="n"
plot(ADOS ~ SocialIQ, data=d2, type="n")
mtext("NonVerbalIQ + VerbalIQ = 0")
lines(S.seq, mu.mean)
shade(mu.PI, S.seq)
shade(S.PI, S.seq)

#dens plot simulated data from model on raw data

sim.m6 = sim(m6, data = d2, n=1e4)
dens(sim.m6, col = "red", xlim = c(-15, 35), ylim = c(0,0.12),  xlab = "ADOS")
par(new=TRUE)
dens(d2$ADOS, xlim = c(-15, 35), ylim = c(0,0.12), xlab = "ADOS")
title("Model 6")


```

4. Let's discuss contents:
4.1. You have three scores for IQ, do they show shared variance? Is that the same in TD and ASD? What does that tell us about IQ?
4.2. You have explored the relation between IQ and ADOS. How do you explain that relation from a cognitive perspective? N.B. You can present alternative hypotheses.
```{r}
# correlation for ASD children
cor(d1$NonVerbalIQ, d1$SocialIQ)
cor(d1$SocialIQ, d1$VerbalIQ)
cor(d1$VerbalIQ, d1$NonVerbalIQ)

#correlation between variable for ASD+TD children
cor(d2$NonVerbalIQ, d2$SocialIQ)
cor(d2$SocialIQ, d2$VerbalIQ)
cor(d2$VerbalIQ, d2$NonVerbalIQ)

d3 = subset(data, c(Visit ==1 & ASD == 0))
d3=d3[complete.cases(d3),]

# correlation for TD children
cor(d3$NonVerbalIQ, d3$SocialIQ)
cor(d3$SocialIQ, d3$VerbalIQ)
cor(d3$VerbalIQ, d3$NonVerbalIQ)
```

5. Bonus questions: Including measurement errors. 
5.1. Let's assume that ADOS has a measurement error of 1. How would you model that in any of the previous models? 
5.2. We know that IQ has an estimated measurement error of 2.12. How would you include that? 


























